\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\lstset{frame=single,backgroundcolor=\color{lightgray},numbers=left,breaklines=true,basicstyle=\ttfamily}


\title{TD Bioinfo}
\author{Pierre Lindenbaum - Institut du Thorax. Nantes. France}
\date{\today}

\begin{document}
\maketitle
\begin{abstract}
Here are few questions to prepare the courses \textbf{"Advanced NCBI"} and \textbf{"Next-Generation-Sequencing"}.\\
The aim of those exercices is to assure that you have a basic knowledge of the tools and vocabulary that will be used during the courses.\\
Only basic linux commands are required here, no external program should be installed.
\end{abstract}

\section{Linux}
Briefly describe the usage of the following linux commands :
\begin{itemize}
\item cd
\item ls
\item pwd
\item ls
\item cp
\item rm
\item mv
\item cat
\item more
\item grep
\item sort
\item uniq
\item paste
\item join
\item tr
\item head
\item tail
\item mkdir
\item awk
\end{itemize}
\noindent
What is 	\textbf{stdout} ? \\
What is 	\textbf{stderr} ?\\
What is 	\textbf{stdin} ?\\
How do you 	\textbf{redirect} the ouput of a program to a file ?\\
How do you 	\textbf{append} the ouput of a program to a file ?\\
\\
In the HTTP protocol, what are the POST and GET methods ?\\
What is the JSON format ?\\
What is a XML well-formed document ?\\
What is a XML DTD ?\\
What is a XML Schema ?\\
What is a XML valid document ?\\

%
% "Nicotiana tabacum"[ORGN] 100:150[SLEN] trnH[GENE] variation[FKEY]
% (Science[JOUR] AND COL1A1[GENE] AND Asara[AUTH]) NOT ("Mammut americanum" [ORGN] OR "Brachylophosaurus canadensis"[ORGN])

\section{General Bioinformatics}
\begin{itemize}
\item What is a 'SNP' ? 
\item What is a 'Genome build' ?
\item What's the approximate size of the human genome ?
\item How many chromosomes is there in the human genome ?
\item What the approximate size of the longest human chromosome ?
\item What's the difference between a binary file and flat file ?
\item How many bytes do you need to store the \textbf{length} of the human genome ?
\item In a sequence containing only 'A', 'T', 'G' and 'C' , how many bases can you store in one byte ?
\item Cite some advantages/inconvenients of storing data in a tab delimited format vs using a structured format (e.g: XML, JSON, ASN.1)
\item Cite some advantages/inconvenients of storing data of using a graphical interface (like Miscrosoft Excel) vs using the command line.
\item Find the following nucleotide entry in the NCBI:\emph{ a Sequence of "Nicotiana tabacum" for the gene 'trnH' with a length comprised between 100 and 150 nucleotides and having a feature "variation"}.
\item Find the following protein entry in the NCBI:\emph{ a sequence for the gene COL1A1 published by Dr. Asara in the "Science" journal,  that is not a sequence of "Mammut americanum" nor "Brachylophosaurus canadensis"}.
\end{itemize}

\section{Bash}
\noindent
What's the purpose of the following bash command line ? ( \url{https://gist.github.com/lindenb/65e98e5752ea26eb9868} )

\begin{lstlisting}[language=bash]
$ curl  "http://hgdownload.cse.ucsc.edu/goldenPath/hg19/chromosomes/chrM.fa.gz" |\
gunzip -c |\
tail -n +2 |\
tr "[:lower:]" "[:upper:]" |\
tr -d '\n' |\
tr "AC" "TG" |\
sed 's/\(.\)/\1#/g' |\
tr "#" "\n" |\
LC_ALL=C sort |\
uniq -c |\
sed 's/^[ ]*//' |\
cut -d ' ' -f1 |\
tr "\n" " " |\
awk '{printf("%f %%\n",$1/($1+$2));}' > result.txt
\end{lstlisting}

\section{Makefile}
\noindent
(wikipedia:) \emph{"In software development, Make is a utility that automatically builds executable programs and libraries from source code by reading files called makefiles which specify how to derive the target program"}.\\You'll find many simple tutorials for \textbf{make} on the web.\\
The following \textbf{Makefile} creates a tab-delimited file 'gene\_ontology\_map.txt' containing the following columns:
\begin{itemize}
\item gene-name(hgnc\_symbol)
\item chromosome
\item start\_position
\item end\_position
\item gene-ontology term.
\end{itemize}
Complete the \textbf{Makefile} by replacing the ???????????? with the correct pipeline ( \url{https://gist.github.com/lindenb/65e98e5752ea26eb9868} ).

\begin{lstlisting}[language=make]
.PHONY: all clean

#
# this is the top default target
#
all: gene_ontology_map.txt

# using the 'join' command
# join gene_and_position.bed and gene_and_go.txt to create 
# a file with the following columns:  
# hgnc_symbol, chromosome, start_position, end_position and GO-ID
gene_ontology_map.txt: gene_and_position.tsv gene_and_go.txt
	join ??????????????????????????????????

# using a web browser: go on biomart http://www.ensembl.org/biomart/martview/
# select 'ensembl Gene'
# select 'dataset: homo sapiens'
# select the following attributes : 
#	Chromosome Name
#	HGNC symbol 
#	Gene Start (bp)
#	Gene End (bp)
# 'show query in XML format' and use it in the curl command below
# keep the lines having a HGNC symbol
# sort on HGNC symbol 
gene_and_position.tsv : 
	curl --form-string query='<?xml version="1.0" ??????????????????????????????????
#
# remove the headers from gene_association.goa_human
# keep the columns DB_Object_Symbol (3) and GO_ID (5)
# and sort on DB_Object_Symbol (=HGNC symbol )
# and remove the duplicated lines
gene_and_go.txt : gene_association.goa_human
	grep -v '^!' $<  | ??????????????????????????????????


#download GO-association for human
#format described in http://www.geneontology.org/GO.format.gaf-1_0.shtml
gene_association.goa_human:
	curl -s "http://cvsweb.geneontology.org/cgi-bin/cvsweb.cgi/go/gene-associations/$@.gz?rev=HEAD" |\
		gunzip -c > $@

#
# cleanup everything
#
clean:
	rm -f gene_association.goa_human gene_and_go.txt  \
		gene_and_position.tsv \
		gene_ontology_map.txt
\end{lstlisting}


\section{The Human Genome}

The chromosomes for the latest Human build are available at: \url{http://hgdownload.cse.ucsc.edu/goldenPath/hg19/chromosomes/}. Answer the following question using a linux command line:
\begin{itemize}
\item What's the length of the chr22 ?
\item Look at the 100000 first lines of the chr22. Explain what you see.
\item Get a count of each base in the chr22
\item Using the linux commands 'curl', 'gunzip', 'tr' and 'rev', get the reverse complement of the chromosome chrM.
\end{itemize}


\section{Using XSLT}
Many NCBI web-services produce a XML document. For example, the following URL is a XML-based list to the databases available from the  \textbf{NCBI}: \url{http://eutils.ncbi.nlm.nih.gov/entrez/eutils/einfo.fcgi}.\\
\begin{lstlisting}[language=xml]
<!DOCTYPE eInfoResult PUBLIC "-//NLM//DTD eInfoResult, 11 May 2002//EN" "http://www.ncbi.nlm.nih.gov/entrez/query/DTD/eInfo_020511.dtd">
<eInfoResult>
  <DbList>
    <DbName>pubmed</DbName>
    <DbName>protein</DbName>
    <DbName>nuccore</DbName>
    <DbName>nucleotide</DbName>
    <DbName>nucgss</DbName>
    <DbName>nucest</DbName>
    <DbName>structure</DbName>
    (...)
\end{lstlisting}
\textbf{XSLT} is a XML specification used to transform a XML to another type of document (XML, HTML or text).
As an example, the XSLT stylesheet \textbf{einfo2html.xsl} available at \url{https://gist.github.com/lindenb/65e98e5752ea26eb9868} transforms  \url{http://eutils.ncbi.nlm.nih.gov/entrez/eutils/einfo.fcgi} to a HTML document. The ouput would start with:\\
\begin{lstlisting}[language=html]
<html><body><ul>
<li><a>pubmed</a></li>
<li><a>protein</a></li>
<li><a>nuccore</a></li>
<li><a>nucleotide</a></li>
<li><a>nucgss</a></li>
<li><a>nucest</a></li>
<li><a>structure</a></li>
<li><a>genome</a></li>
<li><a>assembly</a></li>
    (...)
\end{lstlisting} 
\noindent
Using the command line XSLT processor \textbf{xsltproc}, generate this HTML and visualize it in a web browser.\\
The \textbf{href} attribute is missing in the \textless{}a/\textgreater anchors. For example, the href for 'pubmed' would be :

\begin{lstlisting}[language=html]
<html><body><ul>
<li><a href="http://eutils.ncbi.nlm.nih.gov/entrez/eutils/einfo.fcgi?db=pubmed">pubmed</a></li>
    (...)
\end{lstlisting} 
\noindent
Modify the XSLT stylesheet: using the XSLT element \textless{}xsl:attribute/\textgreater add the missing XML attributes. For example see: \url{http://stackoverflow.com/questions/3321119}.


\end{document}

